---
title: "decipher-algo"
author: "Hugo Englund (rickyhugo)"
date: "9/9/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: 'Decipher Encrypted Messages'
output: github_document
---
#### *Coded by Hugo Englund [(rickyhugo@github)](https://github.com/rickyhugo)*

## Introduction
In this notebook, I will perform a simple variation of Markov Chain Monte Carlo (MCMC) simulation to decipher a hidden message -- a substitution cipher. Since the natural language is assumed to follow certain rules and patterns, e.g., the probability of the letter 'b' to precede 'a' is greater than the one for 'x'. Hence, a sequence of letters, a sentence, should satisfy the first order Markov properties. Therefore, for a substitution cipher, we would like to find the unknown function, the key, $f$, such that $$f: \{\text{code space}\} \longrightarrow \{\text{usual alphabet}\}.$$ With this assumption and with inspiration from the ideas presented in [The Markov Chain Monte Carlo Revolution](https://statweb.stanford.edu/~cgates/PERSI/papers/MCMCRev.pdf), I have successfully deciphered a hidden message from my university lecturer. 


### Fecth transition matrix
Firstly, we have to fetch the transition probability matrix. Usually, one have to estimate this from scratch, but my lecturer had estimated this matrix for the particular message he handed out.
```{r}
# fetch and format transition matrix
trans = read.table('tran_prob.txt')
colnames(trans) = c(toupper(letters), " ")
rownames(trans) = c(toupper(letters), " ")
head(trans)
```

Just for fun, we can visualize the transition matrix to get a grasp of what we are working with.
```{r}
library(ggplot2)
library(reshape2)

# convert transition matrix from wide to long format (required for ggplot2)
trans_l = melt(as.matrix(trans))

ggplot(trans_l, aes(Var2, Var1)) + 
    geom_tile(aes(fill=value)) +
    scale_fill_gradient(low="white", high="brown", limits=c(0,1)) +
    labs(x="Conditional second letter", y="Initial letter", fill="Probability") +
    scale_y_discrete(limits = rev(levels(trans_l$Var1))) +
    coord_equal()
```

### Log-likelihood calculation
Moving on, we need to keep track of the likelihood for the different keys that will be generated. Since we work with quite small probabilities, I decided to us the log-likelihood. This simplifies our calculations and can also prevent zero division errors.
```{r}
loglik = function(trans, txt) {
    # compute log-likelihood given transition probabilities
    dim = nchar(txt) - 1
    res = vector(mode="double", length=dim)
    for (i in 1:dim) {
        idx = strsplit(substring(txt, i, i+1), split="")[[1]]
        res[i] = log(trans[idx[1], idx[2]])
    }
    return(sum(res))
}
```

### Text decoding
Next, we need to "decode" the cipher as we generate new keys in order to compute the log-likelihood and compare to the previous key.
```{r}
transform_txt = function(base_key, key, txt) {
    # deciphers text with the current key
    transformed = txt
    for (i in 1:nchar(txt)) {
        substring(transformed, i, i) = base_key[key==substring(txt, i, i)]
    }
    return(transformed)
}
```

### Decipher algorithm
Now, we are able to compute the log-likelihood and to "decode" ciphers. Then, we can implement the simple algorithm:

1. Start with a preliminary guess, say $f_0$.
2. Compute the log-likelihood, $\mathcal{L}(f_0)$.
3. For $t = 1,2,3,\ldots$, change to $f_t$ by making a random transposition of the values $f_{t-1}$ assigns to two symbols, i.e., switch places of two letters in the current key.
4. Compute $\mathcal{L}(f_t)$; If this is greater than $\mathcal{L}(f_{t-1})$, accept $f_t$.
5. If not, generate a $u \sim Bernoulli\left(\frac{\mathcal{L}(f_t)}{\mathcal{L}(f_{t-1})}\right)$. If $u = 1$, go on with $f_t$, otherwise keep $f_{t-1}$.

We do the fifth step to prevent getting stuck in a local maxima and then miss the optimal key. 
Let's code a little!

        
```{r}
decipher = function(txt, trans, iter_lim=1e4, verbosity=500) {
    # base key for text transformation
    base_key = c(LETTERS, " ")
    
    # initial key
    init_key = base_key[sample(1:27)]
    init_txt = transform_txt(base_key, init_key, txt)
    init_loglik = loglik(trans, init_txt)
    
    prev_key = init_key
    prev_loglik = init_loglik
    
    # init counters
    max_loglik = -1e5
    iter = 0
    suc_cnt = 0
    while (iter <= iter_lim) {
        # next key - randomly switch two letters
        idx = sample(1:length(base_key), 2)
        next_key = replace(prev_key, idx, prev_key[c(idx[2], idx[1])])
        next_txt = transform_txt(base_key, next_key, txt)
        next_loglik = loglik(trans, next_txt)
        
        # compare to previous key
        if (next_loglik > prev_loglik) {
            # update when improved
            prev_loglik = next_loglik
            prev_key = next_key
        } else {
            p = exp(next_loglik - prev_loglik)
            if (rbinom(1, 1, p) == 1) {
                # update randomly
                prev_loglik = next_loglik
                prev_key = next_key
            }
        }
        
        # check for improvements
        if (next_loglik > max_loglik) {
            suc_cnt = suc_cnt + 1
            max_loglik = next_loglik
            max_txt = next_txt
            max_iter = iter
            max_improve = suc_cnt
        }
        
        # print current state
        if (iter %% verbosity == 0) {
            cat(
                "Iteration:", iter, "\n",
                "Succesful changes:", suc_cnt, "\n\n",
                "Message:\n", next_txt, "\n\n",
                "Current log-likelihood:", next_loglik, "\n",
                "Previous log-likelihood:", prev_loglik, "\n",
                "Maximum log-likelihood:", max_loglik, "\n",
                "----------------------------------------------------\n\n"
            )
        }
        # update total iterations
        iter = iter + 1
    }
    # print "optimal" state
    cat(
        "Obtained after iteration:", max_iter, "\n",
        "Succesful changes:", max_improve, "\n\n",
        "Deciphred message:\n", max_txt, "\n\n", 
        "Maximum log-likelihood:", max_loglik
    )
}
```

### Time to dechiper!
Let's find out what the hidden message really is!
```{r}
set.seed(20210909) # for reproducibility 
msg = "empenftpenjcjeumdfsmjbmusnmerbsmbsptnfjuhmsimemkstnmdjrrmutltvmgtmemcspkrtntmpenftpenjcjeumysmws mxusdmdfjcfmpenftpenjcjeumrtinmnfjbmieps bma sntmisvm bmkrtebtmbtuymws vmeubdtvmdjnfmws vmvmcsytbmgecxmnsmptmebmbssumebmksbbjgrtmnftmijvbnmsutmdjrrmdjumemivttmr ucf"

decipher(toupper(msg), trans)
```